\chapter{Customisation: A Short Tour}\label{ch:customize1}

We are about to dive into the details of using \biblatex. As we go
through them, we are going to be thinking a bit about
customisation.

\textsf{Biblatex} is highly customisable. That, in the end, is its
point. At one extreme, if you are setting out to write a bibliography
style for general use you might be customising a great deal. But at
that point it gets very complicated. If you are trying to do anything
that difficult you will really want to understand the internals of
\biblatex\ --- which will involve reading the manual
carefully,\intref{\emph{Manual} \S~4} and also looking at some of the
source code.\sidenote{Especially \texttt{standard.def} and the
  \texttt{.bbx} and \texttt{.cbx} files for the various standard
  files.} This guide is not trying to teach you how to do that. But
there are lots of customisations that are well within the reach of the
`ordinary' user, and it makes sense to discuss them. This chapter
provides a general introduction. Further details are given throughout
the guide in the places where they make most sense, and there is a
chapter\intref{Chapter \ref{ch:recipes}} that deals specifically with
`recipes' for various commonly-needed changes.

In order to keep things simple, I am going quite often to explain
\emph{how} to obtain certain commonly needed changes, without
necessarily explaining exactly why it works---what internal mechanism
is in play.

However, to avoid frustration, you may need to understand some of the
common ways in which changes are made, so that you can include the
right commands in the right way. The final section of this chapter
explains the bare minimum that you need to know.

Finally, do bear in mind that, with \biblatex\ there are often several
ways to achieve a particular result. You may find other suggestions
elsewhere about how to get \biblatex\ to do something, and they may
well be just as good (or better) than the ways I have
suggested. Generally, I've tried to keep things as simple as possible.

\clearpage
\section{The common methods}

\indexstart{customisation!general techniques}
\index{Biblatex!package options|see{options}}
\index{options}
\paragraph{Using an option.}\intref{\emph{Manual} \S~3.1}
Many `customisations' can be achieved simply by setting one of
\biblatex's package options. This is done when you load \biblatex. For
instance, to change the way the bibliography is sorted, you might set
the \verb|sorting| option to \verb|none|, by loading
\begin{verbatim}
\usepackage[style=numeric,
            sorting=none]{biblatex}
\end{verbatim}

Apart from the package itself, some commands also have options. For
instance, there are a number of options that can be given to the
\cs{printbibliography}\intref{See chapter~\ref{ch:customize1}}
command to affect how the bibliography is printed.

\paragraph{Redefining a command.} It's quite common for a change to be
made by redefining a command. For instance, in many cases \biblatex\
uses a command to insert a piece of punctuation, so that if you want
to change the punctuation,\intref{See p~\pageref{sec:punctuation}}
you can redefine the command to produce a different punctuation mark.

For instance, there's a command called \cs{multicitedelim} which
inserts the punctuation between multiple citations. Styles set it
differently, but it's usually a comma or a semicolon. Suppose, for
some really odd reason we wanted to make it `AND'. To do this, you use
\cs{renewcommand}.
\begin{verbatim}
\renewcommand{\multicitedelim}{\addspace AND\space}
\end{verbatim}

\index{punctuation}
\paragraph{Punctuation.} At this point, one feature of
bibliography-bashing is worth attention. If you do much of it, you
will find yourself spending quite a bit of time fiddling with
punctuation. It's not easy, but \biblatex\ has some rather
sophisticated ways of helping you.

When setting punctuation in \biblatex\ commands, it's often best
\emph{not} to use punctuation marks directly: for although they will
work, they don't `play' nicely with some of the clever tricks
\biblatex\ uses to keep punctuation straight. Instead, when engaged in
customisation, use the various commands that \biblatex\ offers.

\csindex{adddot}\csindex{addperiod}\csindex{addcomma}\csindex{addcolon}\csindex{addsemicolon}
\csindex{addspace}\csindex{addnbspace}\csindex{addexclam}\csindex{addquestion}
\index{punctuation!period}\index{punctuation!comma}\index{punctuation!semicolon}\index{punctuation!colon}
\index{spaces}\index{punctuation!exclamation mark}\index{punctuation!question mark}
\begin{margintable}
\begin{tabular}{lll}
\toprule
\cs{adddot}       &  \textbf{.} & (abbreviation dot) \\
\cs{addperiod}    & \textbf{.}  & (regular period) \\
\cs{addcomma}     & \textbf{,}  \\
\cs{addcolon}     & \textbf{:}  \\
\cs{addsemicolon} & \textbf{;}  \\
\cs{addexclam} 	  & \textbf{!}  \\
\cs{addquestion}  & \textbf{?}  \\
\cs{addslash}     & \textbf{\slash} \\
\cs{addspace}     &             & (regular space) \\
\cs{addnbspace}   &             & (unbreakable space) \\
\bottomrule
\end{tabular}
\vspace{3pt}
\caption{The \textbackslash add... commands\label{addcommands}}
\end{margintable}

\index{spaces}
\intref[3ex]{For spacing commands see the \emph{Manual} \S~4.7.4}
These (see table \ref{addcommands}) all take the form
\cs{add...}. Where they often have the edge on simply using
punctuation marks directly, is that they are context sensitive: they
will not add marks if it is inappropriate (for instance, they won't
add a comma after a question-mark), and they will remove unnecessary
white space before the mark. The best way to think of them is as
`context-sensitive' punctuation.

\index{strings!commonly used}
\csindex{DefineBibliographyStrings}
\paragraph{Bibliography strings.} Apart from the bits
of\intref{\emph{Manual} \S~4.9.1} bibliographical information
assembled from the database, all sorts of bits of text get printed in
a bibliography: `ed.', `vol.', `pp.'  and so on. These are handled by
\biblatex\ using bibliography strings. One quite common customisation
is to change these. This is done using the command
\begin{quotation}
\ttfamily
\cs{DefineBibliographyStrings}%
  \{\angled{language}\}\\%
  \quad\{\angled{bibstring} = \angled{definition},\\
  \quad \angled{bibstring} = \angled{definition} \}
\end{quotation} For
instance, if we decided that instead of printing `edited by' or `ed
by', \biblatex\ would produce `conjured by' we could alter the
\verb|byeditor| string:
\begin{verbatim}
\DefineBibliographyStrings{english}% or your language
    { byeditor = {conjured by}, }
\end{verbatim}

\index{formatting!field formats}
\csindex{DeclareFieldFormat}
\csindex{DeclareFieldFormat}
\paragraph{Field formats.} Internally, when it prints some data from
the \verb|.bib| file, \biblatex\ passes the data through a filter that
is defined for that field and entry-type. Although there are some
extra complexities about fields that hold names or lists, the basic
filter is set up using a command \cs{DeclareFieldFormat}.\manref{\S~4.4.2} The basic
structure of this is
\begin{pseudoverb}
\cs{DeclareFieldFormat}[\angled{entrytype}]\{\angled{field}\}\{\angled{format}\}
\end{pseudoverb}
The \angled{format} should just be the definition of a command which
accepts a single input (\texttt{\#1}). At its simplest, it
might simply print this:
\begin{pseudoverb}
\cs{DeclareFieldFormat}\{title\}\{\#1\}
\end{pseudoverb}
would just print the field, whereas
\begin{pseudoverb}
\cs{DeclareFieldFormat}\{title\}\{***\#1***\}
\end{pseudoverb}
would print the title surrounded by three asterisks, or
\begin{pseudoverb}
\cs{DeclareFieldFormat}\{title\}\{\cs{textsc}\{\#1\}\}
\end{pseudoverb}
would print the title in small capitals.

\csindex{DeclareListFormat}\csindex{DeclareNameFormat}
Because of the way that \biblatex\ works there are also formatting
directives for lists and
names.\marginnote{\cs{DeclareListFormat}\\\cs{DeclareNameFormat}}
These are more complex; ordinary users are probably best to leave them
alone; but they may be encountered from time to time in particular recipes.

\index{customisation!hooks}
\paragraph{Hooks.} There\intref{\emph{Manual} \S~4.10.6} are a number
of `hooks' offered by \biblatex\ at which it is made relatively easy
to insert user-defined commands which might do something useful. For
instance \cs{AtEveryCitekey}\csindex{AtEveryCitekey} allows one to
execute some code whenever a source is about to be cited, at a time
when its data is available but has not yet been handled;
\cs{AtEveryBibitem}\csindex{Ateverybibitem} offers a similar facility
in relation to the printing of bibliography items. Particularly
useful, sometimes, is the hook \cs{AtDataInput},\csindex{AtDataInput}
which is executed when data is first `read in' to become available for
citation. From time to time you may be encouraged to use these hooks
for various purposes.

\index{sourcemap directive}
\paragraph{Source mapping.} In fact,\intref{\emph{Manual} \S~4.5.2}
\biblatex\ can `get at' the data even before it is read in---indeed,
while it is being processed by \package{Biber}. The facilities
providing for source mapping enable one to play various useful tricks
with the \verb|.bib| file. We've already seen an example of this on
page \pageref{datamap}.

\paragraph{Deeper and darker.}
All these techniques (options, redefinition of `hook' commands, and
the redefinition of bibliography strings) are fairly simple, and
intended for regular use. Some customisations take us deeper: into the
internal workings of \biblatex. These things include the redefinition
of citation commands, the redefinition of things called bibmacros, and
ultimately the rewriting of bibliography drivers.
\indexstop{customisation!general techniques}

\section{Where to put customisations}

\indexstart{customisation!file locations}
This all depends on whether your customisations are `one off' or
intended to be used in all your documents.
\begin{itemize}
\item For one off customisations:
\begin{itemize}
\item customisation by options should go in the options that you list
  when loading \biblatex.\index{options}
\item most other customisations should go in the document preamble, after
  \biblatex\ has been loaded but before |\begin{document}|. In a \emph{very
  few} cases it's better to put them at some other point. Where that is so
  I specifically mention it.\index{customisation!in preamble}
\end{itemize}
\item If you want a standard set of options to apply to all your
  documents, put them in a file called |biblatex.cfg|, which should be
  put in your \emph{local} \textsc{texmf} tree at
  |/tex/latex/biblatex|. If there is such a file it will be loaded
  after \biblatex. It can be used to redefine commands and to set
  (most, but not all) bibliography options, using
  \cs{ExecuteBibliographyOptions}\braced{\angled{options}}. You
  cannot, however, set the bibliography style there: that at least you
  must do when you load \biblatex.\index{customisation!biblatex.cfg@\texttt{biblatex.cfg}}\index{files!biblatex.cfg@\texttt{biblatex.cfg}}\csindex{ExecuteBibliographyOptions}
\item If\intref{\emph{Manual} \S~4.2} you get to the point that you
  are developing a comprehensive style, you will want to write |.bbx|,
  |.lbx| and |.cbx| files which can be loaded as a style. That,
  however, is beyond the scope of this document.
\end{itemize}
\indexstop{customisation!file locations}

%%% Local Variables:
%%% coding: utf-8
%%% mode: LaTeX
%%% TeX-master: "biblatex-tutorial"
%%% End: