\maketitle

\strut\vspace{10cm}

Copyright \textcopyright\ Paul Stanley \releasedate. The moral right of the author is asserted.

\vspace{2pc}

{\small This handbook is released under a CC Attribution-ShareAlike
  license CC BY-SA:
  \url{http://creativecommons.org/licenses/by-sa/3.0/}. In summary,
  you are free to
\begin{itemize}
\item Share --- copy and redistribute the material in any medium or
  format
\item Adapt --- remix, transform, and build upon the material for any
  purpose, even commercially.
\end{itemize}
The licensor cannot revoke these freedoms as long as you follow the
license terms.

Under the following terms:
\begin{itemize}
\item Attribution --- You must give appropriate credit, provide a link
  to the license, and indicate if changes were made. You may do so in
  any reasonable manner, but not in any way that suggests the licensor
  endorses you or your use.
\item ShareAlike --- If you remix, transform, or build upon the
  material, you must distribute your contributions under the same
  license as the original.
\end{itemize}
}

\vspace{2pc} Typeset using \LaTeX\ in the \package{tufte-book} class,
using Palatino, Helvetica, and Bera Mono fonts.  \cleardoublepage

\tableofcontents

\chapter{Preface}

The \biblatex\ package is a tour de force by its originator (Philip
Lehmann) and its current maintainer(s) (Phil Kime --- who is also
responsible for \package{Biber} --- assisted by others). It is
powerful. But with power comes complexity. The manual is a mine of
information, but sometimes rather overwhelming.

My aim here has been to write something that is a bit more than a mere
introduction, but certainly not a systematic manual. That, after all,
\biblatex\ already has. It is supposed to be, above all, practical:
focussed on explaining not how \biblatex\ works, or exploring all its
possible options and wrinkles, but trying to show how ordinary users
can use it to accomplish reasonably `standard' tasks. It is not
intended to replace the manual; indeed, I have assumed that the reader
will have that to hand. It is not intended as an advanced book for
bibliography-style writers. It is aimed at the ordinary user, who is
looking for practical advice about everyday issues. I hope it will be
useful.

It is a pleasure to acknowledge the help of all those, particularly
those who participate in \TeX-StackExchange, who have directly or
indirectly helped in so many ways.

At the time of writing, the version of \biblatex\ on my system is 3.7.

\hfill\smallcaps{PMS}

\hfill{London, \prefacedate}

%%% Local Variables:
%%% coding: utf-8
%%% mode: LaTeX
%%% TeX-master: "biblatex-tutorial"
%%% End:
